# Data handling with R

<!-- (Nicolas) Non-spatial data handling (30 min) -->
## Data types

- numeric:
 - `2`
 - `12.125`
- character strings: `'a'`, `"word"`
- logical:
  -`TRUE`
  -`FALSE`
  -`NULL` (sort of)

> Please note that logicals are in capital letters.

- vectors:
  - `c(12, 15, 35698)`
  - `c('FOSS4G 2022', 'Firenze')`
- list: `list(1, 45, 12.0, "toto")`
- matrices: `matrix(0:9, 3,3)`
- dataframe (df): `data.frame(x = 1:3, y = c('a', 'b', 'c'))`
- constants:
  - `letters[3:4]`
  - `LETTERS[12:26]`
  - `pi`

## Affectation

Use of variables to store data in memory.

Use `<-` (or `=`)

Examples:

```{r, eval=FALSE}
a <- c(0, 1, 2) # new integer vector
b <- c('FOSS4G', '2022', 'Firenze') # new character vector
c <- data.frame(
  number = a, 
  strings = b)
d <- list(1, 45, 12.0, "toto") # lists can store different data types
e = matrix(0:9, 3,3)
```

## Not only a calculator

R is shipped with lots of functions:

```{r r-functions, eval=FALSE}
data(<datasetname>)   # load an embedded dataset
head(<objectname>)    # first lines of a dataframe
is.vector(object)     # return TRUE if object is a vector
is.data.frame(object) # return TRUE if object is a data.frame
class(<objectname>)  # Returns the class of an object
unique(<objectname>) # returns unique values
help(<functionname>)  # get help on a function
plot(<objectname>)   # create a graphic from a dataset
```

Exercices : 

- Load the `LakeHuron` dataset using `data()`
- Get help on the `LakeHuron` dataset
- Use `head()` to see its first lines
- Plot `LakeHuron`


## Packages

While R base contains a lot of functions, it can be extended with packages.

> In R, the fundamental unit of shareable code is the package. A package bundles together code, data, documentation, and tests, and is easy to share with others. @wickham2022Packages

You already saw how to install packages with the `install.packages()` function, 
let's see how to load the [dplyr](https://dplyr.tidyverse.org/) 
library we will use to handle the data.

```{r data-handling-libraries}
library(dplyr)
```


## Load data

If the packages provides datasets to work with, you will want to work on your own data.
There is several ways to load data.

R base provides a set of functions for delimited text files.
For example, if we want ot work with the Gapminder dataset from [Software Carpentry's R course](https://github.com/swcarpentry/r-novice-gapminder).

```{r data-handling-download-gapminder}
data_url <- "https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/data/gapminder_data.csv"

# download the data and store it in a variable
gapminder <- read.csv(data_url)
```

We call this dataset from an URL but it can be a path in your system file.

For example, if you have installed the gapminder package as recommended, 
you will find the data set as a TSV file in your R installation.

```{r where-is-gapminder}
data_url <- system.file("extdata/gapminder.tsv",package = "gapminder")
data_url
```

:::: {.infobox .tip data-latex="tip"}

`system.file()` is a function that the path of files in R packages, independently
of the operating system.

::::



```{r load-gapminder-from-package}
gapminder <- read.delim(data_url, sep = "\t")
```


:::: {.infobox .tip data-latex="tip"}

`read.csv()` and similars functions can read delimited text files only.
For other formats, you can use functions from other packages like 
[Haven](https://haven.tidyverse.org) or [readxl](https://readxl.tidyverse.org/).
We will show in [Section](#manipulating-vector-data) how to load geospatial data.

::::

Let's take a look to our data :

```{r data-handling-glimpse}
glimpse(gapminder)
```

This dataset is about life expectancy, population and GDP per capita in world 
countries between `r min(gapminder$year)` and `r max(gapminder$year)`.

You can also use `head()` for the same purpose.

```{r data-handling-head}
head(gapminder)
```

:::: {.infobox .note data-latex="note"}

Hans Rosling made an animated presentation of the data [TED talks](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen)

::::

We can see that it contains several columns (also called *variables*):

```{r data-handling-names}
names(gapminder)
```

## In the beginning was the Verb

`dplyr` provides a lot of functions to handle data: `filter()` to filter the data matching
certains conditions (subset data), `select()` to select columns, `mutate()` to create new variables.
All those functions are verbs and means an action onto the dataset.



## Filter data

If we only want the records from Italy, we can filter using 

```{r data-handling-filter}

filter(gapminder, country == "Italy" )

```

You can use comparison operators like `==`, `>`, `<`, `>=` , etc.
There is also logical operators : `&` (*AND*), `|` (*OR*), `!` (*NOT*) and `xor()`.

So, for example, if we want to subset records for Italy after 2000, we can use filter like this:

```{r, data-handling-subset-italy}
italy_2000 <- filter(gapminder, country == "Italy" & year > 2000 )
italy_2000
```

### Exercice {-}

- Try to subset non-European records

```{r class.source = 'fold-hide'}
filter(gapminder, continent != "Europe")
```

- Try to subset records that are in Oceania or before 2000

```{r class.source = 'fold-hide'}
filter(gapminder, continent == "Oceania" | year < 2000)
```

## Select columns

`select`allows you to keep only the columns you need for your analysis.

Maybe we only want the `country`, `year` and `lifeExp` variables:

```{r data-handling-select}

select(italy_2000, country, year, lifeExp)

```

Or, for example, in the `italy_2000` subset, the `continent` variable does not 
provide useful information anymore, so we want to discard it. 
Please not the use of the `-` symbol before the name of the variable.

```{r data-handling-select2}

select(italy_2000, -continent)

```

So you can select the column you want to keep or the ones you want to remove.

## Create new variables

Let say you want to compute the GDP, in millions, from the population and the GDP per capita.
For that, you can use the `mutate()` function:

```{r, data-handling-mutate}
mutate(gapminder, GPD = (gdpPercap * pop)/ 1000000)
```

It's companion function, `transmute()` does the same thing but only keep the new variables.

```{r, data-handling-transmute}
transmute(gapminder, GPD = (gdpPercap * pop)/ 1000000)
```

If you want to keep variables from the dataset, you can call them in the function call:

```{r, data-handling-transmute2}
transmute(gapminder, country, year, GPD = (gdpPercap * pop)/ 1000000)
```

So in this example, we compute the GPD but we also keep the information about the year and the country.

## Agregate data

Sometimes, we have a too detailled dataset and we want a more broader view of the data
so we want to agregate it.

To do so, `dplyr` comes with a couple of functions:  `group_by()` and `summarise()`.
Like their names says, `group_by` groups records who share the same value in a variable 
and `summarise()` compute the summary of non-grouping variables.

```{r, data-handling-group-by}

summarise(group_by(gapminder, continent))

```


## Join data

## Piping

More information on pipe's operators on this [R-bloggers' blogpost](https://www.r-bloggers.com/2021/05/the-new-r-pipe/)
