# Data handling with R

<!-- (Nicolas) Non-spatial data handling (30 min) -->
## Data types

- numeric:
 - `2`
 - `12.125`
- character strings: `'a'`, `"word"`
- logical:
  -`TRUE`
  -`FALSE`
  -`NULL` (sort of)

Please note that logicals are in capital letters.

- vectors:
  - `c(12, 15, 35698)`
  - `c('FOSS4G 2022', 'Firenze')`
- list: `list(1, 45, 12.0, "toto")`
- matrices: `matrix(0:9, 3,3)`
- dataframe (df): `data.frame(x = 1:3, y = c('a', 'b', 'c'))`
- constants:
  - `letters[3:4]`
  - `LETTERS[12:26]`
  - `pi`

## Affectation

Use of variables to store data in memory.

Use `<-` (or `=`)

Examples:

```{r, eval=FALSE}
a <- c(0:9) # new integer vector
b <- c('FOSS4G 2022', 'Firenze') # new character vector
c <- list(1, 45, 12.0, "toto") # lists can store different data types
d = matrix(0:9, 3,3)
```

## Not only a calculator

R is shipped with lots of functions:

```{r r-functions, eval=FALSE}
data(<datasetname>)   # load an embedded dataset
head(<objectname>)    # first lines of a dataframe
is.vector(object)     # return TRUE if object is a vector
is.data.frame(object) # return TRUE if object is a data.frame
typeof(<objectname>)  # Type of an object
unique(<datasetname>) # returns unique values
help(<functionname>)  # get help on a function
plot(<datasetname>)   # create a graphic from a dataset
```

Exercices : 

- Load the `LakeHuron` dataset using `data()`
- Get help on the `LakeHuron` dataset
- Use `head()` to see its first lines
- Plot `LakeHuron`



