# Making maps in R

<!-- (Jakub) Mapping in R with {tmap} (30 min) -->

This chapter requires the following packages:

```{r}
#| message = FALSE
library(tmap)
library(sf)
library(terra)
```

Additionally, we will use three datasets:

- `nz`: a set of polygons representing 16 regions of New Zealand
- `nz_ports`: a point dataset with locations of New Zealand main ports
- `nz_elev`: an elevation raster data of the New Zealand area

```{r}
library(spData)
data("nz")
nz_ports <- read_sf("data/nz_ports.gpkg")
nz_elev <- rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
```

## Mapping tools in R

R has many packages dedicated to spatial data visualizations. 
They include tools for making both static maps, interactive maps, specific-purpose maps, animations, etc.

Packages for creating static maps include **graphics**, **rasterVis**, **ggplot2**, **ggspatial**, **mapsf**, **tidyterra**, etc.
Interactive maps' packages are, for example, **leaflet**, **mapview**, **mapdeck**.
Specific-purpose mapping can be achieved with **cartogram** (to construct area cartograms), **geofacet** ("geofaceting"), **geogrid** (to turn polygons into regular or hexagonal grids), and **rayshader** (raytracing to produce 2D and 3D data visualizations). 

In this workshop, we will focus on the **tmap** package.
It accepts spatial data in various formats, allows to create static and interactive maps, and makes it possible to create small multiples map and map animations.

## Basic example

The code below shows a basic example of the tmap use:

```{r}
tm_shape(nz) +
  tm_graticules() +
  tm_polygons(col = "Median_income", title = "Median income (USD)") +
  tm_shape(nz_ports) +
  tm_symbols(size = 0.75) +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_layout(bg.color = "lightblue")
```

It combines eight function calls using the `+` operator to create the above map.
Each function can be adjusted using its arguments.

We can divide the above tmap functions into a few groups:

- shapes and layers: `tm_shape()`, `tm_polygons()`, `tm_symbols()`. 
They are used to read the spatial data and specify how it should be presented.
- attribute layers: `tm_graticules()`, `tm_scale_bar()`, `tm_compass()`.
They add additional information.
- other map elements: `tm_layout()`.
It specifies the overall map look, including its background color or title.

We expand the explanation of these functions in the next few sections. 

## Shapes and layers

```{r}
#| echo = FALSE
map_layers <- data.frame(Function = c("`tm_polygons()`", "`tm_symbols()`", 
                                     "`tm_lines()`", "`tm_raster()`",
                                     "`tm_text()`"),
                        Geometry = c("polygons", "points, lines, and polygons",
                                     "lines", "raster", "points, lines, and polygons"))
knitr::kable(map_layers, caption = "Basic map layers")
```

```{r}
tm_shape(nz) +
  tm_polygons()
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "darkblue") #or #00008b
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "MAP_COLORS")
```

```{r}
head(nz)
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income")
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis")
```

`tmaptools::palette_explorer()`

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis")
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis", style = "cont")
```

```{r}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis", style = "cont") +
  tm_shape(nz_ports) +
  tm_symbols()
```

```{r}
tm_shape(nz_elev) +
  tm_raster(palette = "-Spectral", style = "cont",
            title = "Elevation (m asl)") +
  tm_shape(nz_ports) +
  tm_symbols()
```

```{r}
tm <- tm_shape(nz_elev) +
  tm_raster(palette = "-Spectral", style = "cont",
            title = "Elevation (m asl)") +
  tm_shape(nz_ports) +
  tm_symbols()
```

## Attributes layers

```{r}
#| echo = FALSE
attr_layers <- data.frame(Function = c("`tm_graticules()`", "`tm_scale_bar()`",
                                      "`tm_compass()`", "`tm_logo()`", "`tm_credits()`"),
                         Description = c("draws latitude and longitude graticules", 
                                         "adds a scale bar", "adds a compass rose",
                                         "adds a logo", "adds a text annotation"))
knitr::kable(attr_layers, caption = "Basic attributes layers")
```

```{r}
tm +
  tm_graticules() +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_logo("https://foss4g.org/logos/2022-v2.png") +
  tm_credits("N. Roelandt and J. Nowosad")
```

```{r}
tm2 <- tm +
  tm_graticules() +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_logo("https://foss4g.org/logos/2022-v2.png") +
  tm_credits("N. Roelandt and J. Nowosad")
```

## Other map elements

```{r}
tm2 +
  tm_add_legend(type = "symbol", col = "grey", title = "Main ports") +
  tm_layout(main.title = "New Zealand", bg.color = "lightblue")
```

```{r}
tm3 <- tm2 +
  tm_add_legend(type = "symbol", col = "grey", title = "Main ports") +
  tm_layout(main.title = "New Zealand", bg.color = "lightblue")
```

## Interactive mode

```{r}
tmap_mode("view")
tm3
```

```{r}
tm3 + 
  tm_minimap() +
  tm_mouse_coordinates()
```

<!-- , "`tm_basemap()`", "`tm_tiles()`" -->

```{r}
tmap_mode("plot")
tm3
```

## Saving maps

```{r}
#| eval = FALSE
tmap_save(tm3, "my_map.png")
```

```{r}
#| eval = FALSE
tmap_save(tm3, "my_map.svg")
```

```{r}
#| eval = FALSE
tmap_save(tm3, "my_map.html")
```

## What else?

- Facet maps
- Animations
- Mapping applications
- `tmap_tip()`

## More resources

- https://r-tmap.github.io/tmap-book/index.html
- https://geocompr.github.io/post/2019/tmap-grid/
- https://geocompr.github.io/post/2019/tmap-color-scales/

## Exercises

<!-- try saving to different formats -->
<!-- try other palettes -->
<!-- try other styles -->
<!-- try tm_add_legend -->
