# Making maps in R

<!-- (Jakub) Mapping in R with {tmap} (30 min) -->

This chapter requires the following packages:

```{r 04-making-maps-1}
#| message = FALSE
library(tmap)
library(sf)
library(terra)
```

Additionally, we will use three datasets:

- `nz`: a set of polygons representing 16 regions of New Zealand
- `nz_ports`: a point dataset with locations of New Zealand main ports
- `nz_elev`: an elevation raster data of the New Zealand area

```{r 04-making-maps-2}
#| message = FALSE
library(spData)
data("nz")
nz_ports <- read_sf("data/nz_ports.gpkg")
nz_elev <- rast(system.file("raster/nz_elev.tif", package = "spDataLarge"))
```

## Mapping tools in R

R has many packages dedicated to spatial data visualizations. 
They include tools for making both static maps, interactive maps, specific-purpose maps, animations, etc.

Packages for creating static maps include {graphics}, {rasterVis}, {ggplot2}, {ggspatial}, {mapsf}, {**}tidyterra}, etc.
Interactive maps' packages are, for example, {leaflet}, {mapview}, {mapdeck}.
Specific-purpose mapping can be achieved with {cartogram} (to construct area cartograms), {geofacet} ("geofaceting"), {geogrid} (to turn polygons into regular or hexagonal grids), and {rayshader} (raytracing to produce 2D and 3D data visualizations). 

In this workshop, we will focus on the {tmap} package.
It accepts spatial data in various formats, allows to create static and interactive maps, and makes it possible to create small multiples map and map animations.

## Basic example

The code below shows a basic example of the {tmap} use:

```{r 04-making-maps-3}
tm_shape(nz) +
  tm_graticules() +
  tm_polygons(col = "Median_income", title = "Median income (USD)") +
  tm_shape(nz_ports) +
  tm_symbols(size = 0.75) +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_layout(bg.color = "lightblue")
```

It combines eight function calls using the `+` operator to create the above map.
Each function can be adjusted using its arguments.

We can divide the above tmap functions into a few groups:

- shapes and layers: `tm_shape()`, `tm_polygons()`, `tm_symbols()`. 
They are used to read the spatial data and specify how it should be presented.
- attribute layers: `tm_graticules()`, `tm_scale_bar()`, `tm_compass()`.
They add additional information.
- other map elements: `tm_layout()`.
It specifies the overall map look, including its background color or title.

We expand the explanation of these functions in the next few sections. 

## Shapes and layers

A simple instance of a (t)map consists of specifying spatial object with `tm_shape()` (this can be, for example, an {sf} vector or a {terra} raster) and then how this object should be visualized.
For example, the code below takes the `nz` object (sf object with polygons) and plots it as polygons:

```{r 04-making-maps-4}
tm_shape(nz) +
  tm_polygons()
```

Table \@ref(tab:04-making-maps-5) gives a list of basic map layers allowed by {tmap}.
We can use different functions depending on our input spatial data type.

```{r 04-making-maps-5}
#| echo = FALSE
map_layers <- data.frame(Geometry = c("polygons", "points, lines, and polygons",
                                     "lines", "raster", "points, lines, and polygons"),
                         Function = c("`tm_polygons()`", "`tm_symbols()`", 
                                     "`tm_lines()`", "`tm_raster()`",
                                     "`tm_text()`"))
knitr::kable(map_layers, caption = "Basic map layers")
```

As we have seen above, polygons can be visualized with `tm_polygons()`; however, we can also show them using `tm_symbols()` or `tm_text()`:

```{r}
tm_shape(nz) +
  tm_symbols()
```

Each map layer can be adjusted.
For example, `tm_polygons()` can be represented by either:

- one consistent color (e.g., `col = "darkblue"`)
- unique colors for adjacent polygons (`col = "MAP_COLORS"`)
- color representing values of a given variable (`col = Median_income`)

One consistent color may be set with [a color name](https://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) or a hex color code:

```{r 04-making-maps-6}
tm_shape(nz) +
  tm_polygons(col = "darkblue") #or #00008b
```

To use unique colors for adjacent polygons we need to use `col = "MAP_COLORS"`: 

```{r 04-making-maps-7}
tm_shape(nz) +
  tm_polygons(col = "MAP_COLORS")
```

We also can provide a variable name to color each polygon based on its value (note: you can check your variables names with `head(nz)`).
Let's create a choropleth map of the `"Median_income"` variable:

```{r 04-making-maps-9}
tm_shape(nz) +
  tm_polygons(col = "Median_income")
```

Next, we customize the map with additional arguments, such as `title` to change the legend title or `palette` to provide a name of the color palette to use:

```{r 04-making-maps-10}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis")
```

The {tmap} package accepts names of a few dozens of color palettes (note: try `tmaptools::palette_explorer()`), but it is also possible to provide a vector of colors here.

By default, {tmap} behaves differently depending on the input variable type, e.g., uses unique colors for categorical variables and pretty breaks for continuous variables.
We can use [the `style` argument](https://geocompr.github.io/post/2019/tmap-color-scales/) if we want to change the color breaks.
For example, `style = "cont"` creates a continuous color gradient:

```{r 04-making-maps-12}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis", style = "cont")
```

Many spatial objects and their related map layers can be connected with the `+` operator.
Importantly, subsequent map layers are drawn on top of the previous ones:

```{r 04-making-maps-13}
tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis", style = "cont") +
  tm_shape(nz_ports) +
  tm_symbols()
```

The previous examples used spatial vector data.
However, plotting raster data works in the same fashion -- we need to provide a spatial object with `tm_shape()`, and then plot it with `tm_raster()`:

```{r 04-making-maps-14}
#| message = FALSE
tm_shape(nz_elev) +
  tm_raster(title = "Elevation (m asl)",
            palette = "-Spectral", style = "cont") +
  tm_shape(nz_ports) +
  tm_symbols()
```

The tmap functions do not only result in a plot, but their output can be also attach to an R object:

```{r 04-making-maps-15}
tm <- tm_shape(nz) +
  tm_polygons(col = "Median_income", title = "Median income (USD)",
              palette = "viridis", style = "cont") +
  tm_shape(nz_ports) +
  tm_symbols()
```

This is useful when we want to add new layers to our map or save it to a file.

## Attributes layers

Attributes layers allow to draw often used map elements, such as [graticules](https://geocompr.github.io/post/2019/tmap-grid/), scale bars, north arrows, logos, or credits (Table \@ref(tab:04-making-maps-16)).

```{r 04-making-maps-16}
#| echo = FALSE
attr_layers <- data.frame(Description = c("draws latitude and longitude graticules", 
                                         "adds a scale bar", "adds a compass rose",
                                         "adds a logo", "adds a text annotation"),
                            Function = c("`tm_graticules()`", "`tm_scale_bar()`",
                                      "`tm_compass()`", "`tm_logo()`", "`tm_credits()`"))
knitr::kable(attr_layers, caption = "Basic attributes layers")
```

We can add these elements to our previous map with the following code:

```{r 04-making-maps-17}
#| message = FALSE
tm +
  tm_graticules() +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_logo("https://foss4g.org/logos/2022-v2.png") +
  tm_credits("N. Roelandt and J. Nowosad")
```

Each map element can be also customized (e.g., by specifying breaks for the scale bar or a text in the credits), and its location can be set with `position`.

Let's save our new map to the `tm2` object:

```{r 04-making-maps-18}
#| message = FALSE
tm2 <- tm +
  tm_graticules() +
  tm_scale_bar(breaks = c(0, 100, 200)) +
  tm_compass(position = c("right", "top")) +
  tm_logo("https://foss4g.org/logos/2022-v2.png") +
  tm_credits("N. Roelandt and J. Nowosad")
```

## Other map elements

The {tmap} also has some other map elements.
It includes `tm_add_legend()` that allows to add a manual legend by specifying its type, color, and title.
In the example below, we also use `tm_layout()` -- this function specifies the overall map look, including its background color, title, fonts, etc.

```{r 04-making-maps-19}
#| message = FALSE
tm2 +
  tm_add_legend(type = "symbol", col = "grey", title = "Main ports") +
  tm_layout(main.title = "New Zealand", bg.color = "lightblue")
```

Here, we save our new map to the `tm3` object:

```{r 04-making-maps-20}
tm3 <- tm2 +
  tm_add_legend(type = "symbol", col = "grey", title = "Main ports") +
  tm_layout(main.title = "New Zealand", bg.color = "lightblue")
```

## Interactive mode

Each map created with {tmap} can be viewed in either `"plot"` and `"view"` mode: the default `"plot"` mode returns a static map, while the `"view"` mode results in an interactive map.
We can change the mode with `tmap_mode()`:

```{r 04-making-maps-21}
tmap_mode("view")
```

Then we just need to open the map object (or write a (t)map code):

```{r}
#| collapse = TRUE,
#| results = "hold"
tm3
```

As you can see above, the interactive mode has the same map layers, colors, and legend.
However, both modes have their own features.
For example, `tm_compass()`, `tm_logo()`, and `tm_credits()` only work in the static mode.
On the other hand, the interactive mode allows for zooming or panning, and also makes it possible to select and change the background tiles.^[Try `tm_basemap()` and `tm_tiles()` if you want to customize the interactive tiles.] We can also add some map elements available for interactive mode only, such as `tm_minimap()` or `tm_mouse_coordinates()`.

```{r 04-making-maps-22}
#| collapse = TRUE,
#| results = "hold"
tm3 + 
  tm_minimap() +
  tm_mouse_coordinates()
```

To return to the static mode, we need to use `tmap_mode("plot")`:

```{r 04-making-maps-23}
#| message = FALSE
tmap_mode("plot")
tm3
```

## Saving maps

Maps created with {tmap} can be saved to various file formats.
It includes `.png` for raster graphic files, `.svg` for vector graphic files, or even `.html` to save an interactive map.
All of the saving can be done with `tmap_save()`, which accepts the map object and a file path.
It also allows customizing the output map resolution.

```{r 04-making-maps-24}
#| eval = FALSE
tmap_save(tm3, "my_map.png")
tmap_save(tm3, "my_map.svg")
tmap_save(tm3, "my_map.html")
```

:::: {.infobox .tip data-latex="note"}

Note that some {tmap} functions have a `tm_` prefix, while other a `tmap_` prefix. 
This syntax allows to distinct between making maps functions (e.g., `tm_shape()` or `tm_polygons()`) and other functions (e.g., `tmap_save()`).

::::

## What else?

The above examples showed basic (and probably the most often used) features of {tmap}.
However, this package has much more to offer, including:

- Facet maps: using `tm_facets()` to create small multiples map
- Animations: using `tm_facets()` + `tmap_animation()` to create map animations
- Mapping applications: using the {shiny} package with `renderTmap()`, `tmapOutput()`, etc.
- `tmap_tip()`

## More resources

For more resources check [the {tmap} repo](https://github.com/r-tmap/tmap), read [the tmap book (work in progress)](https://r-tmap.github.io/tmap-book) and [the Making maps with R chapter](https://geocompr.robinlovelace.net/adv-map.html).

## Exercises

Read the following datasets:

```{r}
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
zion <- read_sf((system.file("vector/zion.gpkg", package = "spDataLarge")))
zion_points <- read_sf(system.file("vector/zion_points.gpkg", package = "spDataLarge"))
```

E1. Create a simple map using {tmap} with three map layers: (1) `srtm` (colored based on its values), (2) `zion` (as a grey polygon), (3) `zion_points` (as grey symbols).

```{r}
#| include = FALSE
tm_shape(srtm) +
  tm_raster() +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols()
```

E2. Customize the map from E1, for example by: improving the legend title, changing the color palette, etc.

```{r}
#| include = FALSE
tm_shape(srtm) +
  tm_raster(style = "cont", palette = "-Spectral", title = "Elevation (m als)") +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols(shape = 6, col = "black")
```

E3. Add some attribute layers to your map, including the scale bar and north arrow. 
Also, add your name in the bottom left corner of the map.

```{r}
#| include = FALSE
tm_shape(srtm) +
  tm_raster(style = "cont", palette = "-Spectral", title = "Elevation (m als)") +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols(shape = 6, col = "black") +
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 2, 4)) +
  tm_compass(position = c("right", "top")) +
  tm_credits("My Name", position = c("left", "bottom"))
```

E4. Try to adjust the overall map look by, for example, removing the map frame and adding a map title.

```{r}
#| include = FALSE
tm_shape(srtm) +
  tm_raster(style = "cont", palette = "-Spectral", title = "Elevation (m als)") +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols(shape = 6, col = "black") +
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 2, 4)) +
  tm_compass(position = c("right", "top")) +
  tm_credits("My Name", position = c("left", "bottom")) +
  tm_layout(frame = FALSE, main.title = "Zion National Park",
            legend.position = c("left", "center"))
```

E5. Bonus: add two manual legend: one for the `zion` and one for `zion_points`.

```{r}
#| include = FALSE
tm_shape(srtm) +
  tm_raster(style = "cont", palette = "-Spectral", title = "Elevation (m als)") +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols(shape = 6, col = "black") +
  tm_add_legend(type = "symbol", shape = 6, col = "black", labels = "My points") +
  tm_add_legend(type = "fill", alpha = 0.3, labels = "Park's area") +
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 2, 4)) +
  tm_compass(position = c("right", "top")) +
  tm_credits("My Name", position = c("left", "bottom")) +
  tm_layout(frame = FALSE, main.title = "Zion National Park",
            legend.position = c("left", "center"))
```

E6. Try saving your map to different file formats: `.png`, `.pdf`, and `.html`. 
Can you notice any difference between the files?

```{r}
#| include = FALSE
my_map <- tm_shape(srtm) +
  tm_raster(style = "cont", palette = "-Spectral", title = "Elevation (m als)") +
  tm_shape(zion) +
  tm_polygons(alpha = 0.3) +
  tm_shape(zion_points) +
  tm_symbols(shape = 6, col = "black") +
  tm_add_legend(type = "symbol", shape = 6, col = "black", labels = "My points") +
  tm_add_legend(type = "fill", alpha = 0.3, labels = "Park's area") +
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 2, 4)) +
  tm_compass(position = c("right", "top")) +
  tm_credits("My Name", position = c("left", "bottom")) +
  tm_layout(frame = FALSE, main.title = "Zion National Park",
            legend.position = c("left", "center"))
tmap_save(my_map, "my_map2.png")
tmap_save(my_map, "my_map2.pdf")
tmap_save(my_map, "my_map2.html")
```

