# Manipulating vector data {#manipulating-vector-data}

<!-- (Nicolas) Vector data processing (45 min) -->


For this part, we will use mostly `sf` capabilities. 
Most spatial functions (if not all) start with the `st_` (*for spatial type*) 
prefix like in [PostGIS](https://postgis.net/docs/manual-2.5/reference.html).

We will focus on vector data from the `spData` package.

```{r libraries}
library(sf)
library(spData)
library(dplyr)
library(here)
```

We will also use functions from the `dplyr` package, `sf` objects are dataframes
compatibles with the Tidyverse philosophy.

## Read spatial data

`sf` provides the `st_read` and `st_write` functions to access geospatial files. 
They can operate with any [vector driver provided by GDAL](https://gdal.org/drivers/vector/index.html).

We will use the data from the `spData`^[More details on the datasets here :
[https://cran.r-project.org/web/packages/spData/spData.pdf](https://cran.r-project.org/web/packages/spData/spData.pdf)] 
package. Let's see what it contains :

```{r spData_files}
list.files(system.file("shapes", package = "spData"))
```

We will work on cycle hires in London so let's load that data.

### Cycle hire dataset

```{r load_hire_data}
cycle_hire <- st_read(system.file("shapes/cycle_hire.geojson", package="spData"))
```

Here we can see a couple of functions :

* `st_read()` is the reading function from `sf`
* `system.file()` is a function that allow us to look for data in packages, independently of the operating system.

By default, `sf` provides informations when loading the dataset. 
We can see it contains 742 features and 5 fields. 
It is points with lat/lon coordinates (we can see it through information about 
its CRS: `Geodetic CRS:  WGS 84`). 
`sf` also provides the bounding box.

Here is the description of the dataset we can find in the documentation: 

```{block2, type='rmdnote'}
**cycle_hire dataset**
  
~Description:~ Points representing cycle hire points accross London.

~Format:~
  
 *  **id** Id of the hire point
 *  **name** Name of the point
 *  **area** Area they are in
 *  **nbikes** The number of bikes currently parked there  
 *  **nempty** The number of empty places
 *  *geometry* sfc_POINT

~Source~: [cyclehireapp.com/cyclehirelive/cyclehire.csv](cyclehireapp.com/cyclehirelive/cyclehire.csv)
```

We can see how many bike are parked, the count of empty slots but not the total 
amount of bike slots. Let's create a new `slots` column for this with `mutate()`
from `dplyr`.

```{r create_bike_slots}
cycle_hire <- mutate(cycle_hire, slots = nbikes +  nempty)
```

Let's load polygon data, in this case London's boroughs stored in the *lnd* dataset.

### Boroughs of London

This dataset is contained in a R data format so the loading is different.

```{r loading_lnd}
data(lnd) # load the dataset in memory
lnd       # call the dataset to visualise the 10 first features
```

We can see this dataset contains 33 features and 7 fields, in lat/lon coordinates too.
The geometry type is different: MULTIPOLYGON.

```{block2, type='rmdnote'}
**ldn dataset**

The boroughs of London

Description : Polygons representing large administrative zones in London

Format:
  
  *  **NAME** Borough name
  *  **GSS_CODE** Official code
  *  **HECTARES** How many hectares
  *  **NONLD_AREA** Area outside London
  *  **ONS_INNER** Office for national statistics code
  *  **SUB_2009** Empty column
  *  **SUB_2006** Empty column
  *  *geometry* sfc_MULTIPOLYGON

Source : [https://github.com/Robinlovelace/Creating-maps-in-R](https://github.com/Robinlovelace/Creating-maps-in-R)
```

In order to ease spatial calculations, let's reproject them.

## Reprojection {#reprojection}

The [Ordnance Survey National Grid](https://en.wikipedia.org/wiki/Ordnance_Survey_National_Grid) is the official one for Great Britain. Its SRID is **EPSG:27700**.

```{r data_reprojection}
cycle_hire_27700 <- st_transform(cycle_hire, crs = st_crs(27700))

london_27700 <- st_transform(lnd, crs = st_crs(27700))
```

To reproject, we used 2 functions:
- `st_transform()` for the reprojection operation
- `st_crs()` to get CRS definition from an EPSG code

Now, we can create a quick map to visualize our data. 
We can use the `plot()` function to do this. 
This function is part of base R.

```{r quick_plot}
plot(london_27700$geometry) # we just want to plot the geometry column
plot(cycle_hire_27700$geometry, 
 col = "red",  # color
 cex = 0.5,    # size of symbol
 add = TRUE)   # important parameter to create multilayer plots
```


## Joins

We can use two ways to link those datasets together, by attributes, 
they share their area name (`area` and `NAME`) or spatially. 
For the sake of the exercice, let's do both.

### Join by attributes

Let's join them  with a inner join to see how many corresponds.

```{r inner_attribute_join}
inner_join(cycle_hire_27700,
  st_drop_geometry(london_27700), # we don't need the geometry here
  by = c( "area" = "NAME")
)
```


We can see that only 33 features matched. 
That's poor, let's try this spatially.

### Spatial join

For this, we will try to provide a `GSS_CODE` for all  cycle hire points. 
We will regroup the data afterwards.

For this, we will select only the `GSS_CODE` column from `london_27700` with the
`select` function from `dplyr`, the geometry will follow.

```{r spatial_join}
cycle_hire_27700 <- st_join(cycle_hire_27700, select(london_27700, GSS_CODE))
```

Now if we look at our dataset, there is a `GSS_CODE` column.

```{r names_cycle_hire}
names(cycle_hire_27700)
```

How many points doesn't have a GSS_code ?

```{r controle_nas}
filter(cycle_hire_27700, is.na(GSS_CODE))
```

Only one, that's more better than before ! 
I don't know well enough London to fix this. 
But that is not blocking.

Now to paraphrase Anita Graser   : [[@graserAggregate]](https://anitagraser.com/2017/06/08/aggregate-all-the-things-qgis-expression-edition/)

## Aggregation
### Count

```{r aggregation_count}
# remove NAs
cycle_hire_clean <- filter(cycle_hire_27700, !is.na(GSS_CODE))

# let's put geometry aside
cycle_hire_clean <- st_drop_geometry(cycle_hire_clean)

# group data by GSS_CODE
cycle_hire_grouped <- group_by(cycle_hire_clean, GSS_CODE)

# count
cycle_hire_by_area <- tally(cycle_hire_grouped, name = "count", sort= TRUE) # Aggregate
cycle_hire_by_area
```

:::: {.infobox .tip data-latex="tip"}

`tally()` is equivalent to `df %>% summarise(n = n())`

::::

### Sum

```{r aggregation_sum}
# count
cycle_hire_by_area_sum <- summarise(
  cycle_hire_grouped, # we reused grouped data
  sum = sum(nbikes), # sums the number of bikes
  count = n() # count cycle stations
  ) 
cycle_hire_by_area_sum
```

We could have use the base function `aggregate()` which works with `sf` objects.

```{r aggregate_alternative}
aggregate(cycle_hire_27700["nbikes"], 
          by = list(cycle_hire_27700$"GSS_CODE"),
          FUN = sum, 
          na.rm = TRUE)
```

If we want to represents our data with proportional symbols, we might want to have centroids. 
`sf` provides two functions in order to do that:

* `st_centroid()`
* `st_point_on_surface()`

`st_point_on_surface()` provides a point randomly **in** the entry shape. 
That can be useful for irregular shapes where the centroid might be outside the shape.

## Centroids

```{r centroids}
 # only keep useful columns
boroughs <- select(london_27700, NAME, GSS_CODE)

# compute centroids
boroughs_centroids <- st_centroid(boroughs)
```

You can also do buffers and other geometrical operations like [`st_union()`](https://r-spatial.github.io/sf/reference/geos_combine.html) to merge geometries

![Spatial equivalents of logical operators [@lovelace_geocomputation_2019]](images/venn-clip-1.png)


## Geometric binary predicates

`sf` provides numerous geometric binary predicates that can be used with the intersection function.

* st_intersects()
* st_disjoint()
* st_touches()
* st_crosses()
* st_within()
* st_contains()
* st_contains_properly()
* st_overlaps()
* st_equals()
* st_covers()
* st_covered_by()
* st_equals_exact()
* st_is_within_distance()

You can use it alone or with  ̀st_join()`. 

For example, if we want to the cycle hires contained in the borough of Wandsworth, 
we will do like this:

```{r find_E09000032}
Wandsworth <- filter(london_27700, NAME == "Wandsworth")
(Wandsworth_bike_stations <- st_contains(Wandsworth, cycle_hire_27700) )
```

That will return a list of cycle hire points id.

In contrary, if we want to find in which borough the hire point with id 614 is 
we can use the opposite function : `st_within()`.

```{r borough_of_614}
cycle_hire_614 <- filter(cycle_hire_27700, id == "614") 
cycle_hire_614_borough <- st_within(cycle_hire_614, london_27700) # borough at index 22
cycle_hire_614_borough
```

To get the borough data, there is some more work to do.

```{r get_borough22}
london_27700[unlist(cycle_hire_614_borough),]
```

## Saving results


In the first part, we saw that we can read data but we can also write it !

### Writing data

To write data, we will use the `st_write()` function.
It takes the data source name (*dsn*) as mandatory argument, 
`sf` will try to find the good driver from the extension 
(here it is *gpkg* for GeoPackage).

```{block2, type='rmdwarning'}
*st_write()* can't save non geospatial data. So we need to join the data from cycle_hire_by_area_sum to
the boroughs first.
```

As we want to save it to  GeoPackage^[Because [GeoPackage are cool !](http://switchfromshapefile.org/#geopackage)], we also need to provide a layer name : *london_boroughs_27700*. Repeat for all data you want to save.

```{r write_geodata}
st_write(
  obj = left_join(london_27700, cycle_hire_by_area_sum), # object to write
  dsn = here("foss4g_R_workshop.gpkg"), # destination file
  layer = "london_boroughs_27700",   # layer name
  append = FALSE)  # options

st_write(
  left_join(boroughs_centroids , cycle_hire_by_area_sum),
  dsn = here("foss4g_R_workshop.gpkg"),
  layer = "boroughs_centroids_27700", 
  append = FALSE)

st_write(
  obj = left_join(cycle_hire_27700, cycle_hire_by_area_sum),
  dsn = here("foss4g_R_workshop.gpkg"),
  layer = "cycle_hire_27700",
  append = FALSE)
```

```{block2, type='rmdnote'}
We used the `here()` function as it preserve the project file hierarchy. 
It works better in Rstudio but it is still useful with Jupyter notebooks.

The data set where joined by their GSS_CODE. You can specify the "by" statement, but for the sake of readability, it is not show here.

The `append = FALSE` ensure you can write on existing layer, it is optionnal.
```


```{r here}
print(here())  # print the project directory
list.files(here()) # list the files in the project directory
```

### Check data

`sf` provides a `st_layers()` function that is useful to see the content of a dataset.
 
```{r check_gpkg}
st_layers(dsn = here("foss4g_R_workshop.gpkg"))
```
