# Manipulating raster data

<!-- (Jakub) Raster data processing (45 min) -->

```{r 06-manipulating-raster-data-1}
#| message = FALSE
library(tmap)
library(sf)
library(terra)
```

- The terra package contains classes and methods representing raster objects
- It allows raster data to be loaded and saved
- It allows raster algebra and raster processing
- It includes a number of additional functions, e.g., for analysis of terrain characteristics
- It allows you to work on large sets of data
- ?terra-package, https://rspatial.github.io/terra/reference/terra-package.html
- It is a successor of the raster package - http://www.rpubs.com/etiennebr/visualraster


## Raster reading

```{r 06-manipulating-raster-data-2}
srtm_path = system.file("raster/srtm.tif", package = "spDataLarge")
srtm_path
```

**terra** - https://rspatial.github.io/terra/reference/terra-package.html

```{r 06-manipulating-raster-data-3}
srtm = rast(srtm_path)
srtm
```

```{r 06-manipulating-raster-data-4}
plot(srtm)
```

```{r 06-manipulating-raster-data-5}
tm_shape(srtm) +
  tm_graticules() +
  tm_raster(style = "cont", 
            title = "elevation (m a.s.l)",
            palette = "-Spectral") +
  tm_scale_bar(breaks = c(0, 2, 4),
               text.size = 1) +
  tm_credits("Jakub Nowosad, 2021") +
  tm_layout(inner.margins = 0,
    main.title = "Zion National Park")
```


```{r 06-manipulating-raster-data-6}
ncol(srtm)
nrow(srtm)
nlyr(srtm)
res(srtm)
ext(srtm)
```

```{r 06-manipulating-raster-data-7}
nlcd = rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
```

## Map algebra

Used for a various task related to spatial raster data.

It can be divided into four groups:

1. **Local** - per-cell operations
2. **Focal (neighborhood operations)** - most often the output cell value is the result of a 3 x 3 input cell block
3. **Zonal operations** - to summarize raster values for some zones (usually irregular areas)
4. **Global** - to summarize raster values for one or several rasters

### Local operations

- Raster calculator
- Replacing values
- Reclassification
- Operations on many layers (e.g., calculating spectral indices, such as NDVI)

```{r 06-manipulating-raster-data-8}
srtm2 = srtm + 1000
```

```{r 06-manipulating-raster-data-9}
srtm3 = srtm - 1024
```

```{r 06-manipulating-raster-data-10}
srtm4 = srtm - 1837
```

```{r 06-manipulating-raster-data-11, purl=FALSE,echo=FALSE, message=FALSE, out.width="100%", fig.height=3.8, fig.width=12}
library(tmap)
tm1 = tm_shape(srtm) +
  # tm_graticules() +
  tm_raster(style = "cont")
tm2 = tm_shape(srtm2) +
  # tm_graticules() +
  tm_raster(style = "cont")
tm3 = tm_shape(srtm3) +
  # tm_graticules() +
  tm_raster(style = "cont")
tm4 = tm_shape(srtm4) +
  # tm_graticules() +
  tm_raster(style = "cont")
tmap_arrange(tm1, tm2, tm3, tm4, nrow = 1)
```

```{r 06-manipulating-raster-data-12}
srtm_new = srtm
srtm_new[srtm_new < 1500] = NA
```

```{r 06-manipulating-raster-data-13, purl=FALSE,echo=FALSE, message=FALSE, out.width="100%", fig.height=3.8, fig.width=12}
library(tmap)
tm1 = tm_shape(srtm) +
  # tm_graticules() +
  tm_raster(style = "cont")
tm2 = tm_shape(srtm_new) +
  # tm_graticules() +
  tm_raster(style = "cont")
tmap_arrange(tm1, tm2, nrow = 1)
```

```{r 06-manipulating-raster-data-14}
rcl = matrix(c(0, 1500, 1, 1500, 2000, 2, 2000, 9999, 3),
             ncol = 3, byrow = TRUE)
rcl
```

```{r 06-manipulating-raster-data-15}
srtm_recl = classify(srtm, rcl = rcl)
```

```{r 06-manipulating-raster-data-16, purl=FALSE,echo=FALSE}
tmr1 = tm_shape(srtm) +
  # tm_graticules() +
  tm_raster(style = "cont") +
  tm_layout(legend.frame = TRUE)
tmr2 = tm_shape(srtm_recl) +
  # tm_graticules() +
  tm_raster(style = "cat") +
  tm_layout(legend.frame = TRUE)
tmap_arrange(tmr1, tmr2)
```

`?lapp`

```{r 06-manipulating-raster-data-17}
landsat_path = system.file("raster/landsat.tif", package = "spDataLarge")
landsat = rast(landsat_path)
landsat
```

https://bleutner.github.io/RStoolbox/rstbx-docu/spectralIndices.html

$$
\begin{split}
NDVI&= \frac{\text{NIR} - \text{Red}}{\text{NIR} + \text{Red}}\\
\end{split}
$$

```{r 06-manipulating-raster-data-18}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
ndvi = lapp(landsat[[c(4, 3)]],
            fun = ndvi_fun)
```

```{r 06-manipulating-raster-data-19, purl=FALSE,echo=FALSE, message=FALSE, out.width="100%", fig.height=7}
tm_shape(ndvi) +
  tm_raster(style = "cont", title = "NDVI")
```

### Focal operations

`?focal`

```{r 06-manipulating-raster-data-20}
srtm_focal_mean = focal(srtm, 
                   w = matrix(1, nrow = 3, ncol = 3), 
                   fun = mean)
```

```{r 06-manipulating-raster-data-21, purl=FALSE,echo=FALSE}
tmf1 = tm_shape(srtm) +
  # tm_graticules() +
  tm_raster(style = "cont") +
  tm_layout(legend.frame = TRUE)
tmf2 = tm_shape(srtm_focal_mean) +
  # tm_graticules() +
  tm_raster(style = "cont") +
  tm_layout(legend.frame = TRUE)
tmap_arrange(tmf1, tmf2)
```

### Zonal operations

`?zonal`
Also known as zonal statistics. Result - a summary table

```{r 06-manipulating-raster-data-22}
srtm_utm = project(srtm, nlcd, method = "bilinear")
```

```{r 06-manipulating-raster-data-23, purl=FALSE, warning=FALSE, echo=FALSE, message=FALSE}
tm_shape(nlcd) +
  tm_raster(drop.levels = TRUE, legend.show = FALSE)
```

```{r 06-manipulating-raster-data-24}
srtm_zonal = zonal(srtm_utm, nlcd, na.rm = TRUE, fun = "mean")
srtm_zonal
```

### Global operations 

```{r 06-manipulating-raster-data-25}
global(srtm, fun = "mean")
global(srtm, fun = "sd")
```

```{r 06-manipulating-raster-data-26}
freq(nlcd)
```

<!-- - global + freq + areas (cellSize()) -->

## Transformations

### Resampling

```{r 06-manipulating-raster-data-27}
srtm
```

```{r 06-manipulating-raster-data-28}
new_srtm = srtm
res(new_srtm) = 0.001
new_srtm
```

```{r 06-manipulating-raster-data-29}
srtm2 = resample(srtm, new_srtm, method = "bilinear") # method!
srtm2
```

### Reprojecting rasters

```{r 06-manipulating-raster-data-30}
crs(srtm, describe = TRUE)
```

```{r 06-manipulating-raster-data-31}
hist(srtm)
```

```{r 06-manipulating-raster-data-32, purl=FALSE,echo=FALSE}
tm_shape(srtm) + 
  tm_grid() +
  tm_raster(style = "cont",
            legend.show = FALSE)
```

https://projectionwizard.org/

```{r 06-manipulating-raster-data-33}
srtm_utm = project(srtm, "EPSG:32612", method = "bilinear")
crs(srtm_utm, describe = TRUE)
```

```{r 06-manipulating-raster-data-34, purl=FALSE,echo=FALSE, fig.height=7}
tm_shape(srtm) + 
  tm_grid() + 
  tm_raster(style = "cont") + 
  tm_layout(legend.show = FALSE)
```

```{r 06-manipulating-raster-data-35, purl=FALSE,echo=FALSE, fig.height=7, eval=FALSE}
tm_shape(srtm_utm) + 
  tm_grid() + 
  tm_raster(style = "cont") + 
  tm_layout(legend.show = FALSE)
```


## Raster-vector interactions

- Raster cropping and masking
- Raster extraction - by points, lines, and polygons
- Rasterization - points, lines, polygons to rasters
- Vectorization - rasters to polygons or contours

### Raster cropping and masking

```{r 06-manipulating-raster-data-36}
library(sf)
zion = read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
```

```{r 06-manipulating-raster-data-37}
srtm_utm_c = crop(srtm_utm, vect(zion))
```

```{r 06-manipulating-raster-data-38, echo=FALSE, purl=FALSE, message=FALSE, warning=FALSE, fig.width=12, eval=FALSE}
tc1 = tm_shape(srtm_utm) +
  tm_grid() + 
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_borders() +
  tm_layout(inner.margins = 0)
tc2 = tm_shape(srtm_utm_c) +
  tm_grid() + 
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_borders() +
  tm_layout(inner.margins = 0)
tmap_arrange(tc1, tc2)
```

Raster masking is usually done together with cropping.

```{r 06-manipulating-raster-data-39}
srtm_utm_m = mask(srtm_utm_c, vect(zion))
```

```{r 06-manipulating-raster-data-40, echo=FALSE, purl=FALSE, message=FALSE, warning=FALSE, fig.width=12, eval=FALSE}
tc1 = tm_shape(srtm_utm) +
  tm_grid() + 
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_borders() +
  tm_layout(inner.margins = 0)
tc2 = tm_shape(srtm_utm_c) +
  tm_grid() + 
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_borders() +
  tm_layout(inner.margins = 0)
tc3 = tm_shape(srtm_utm_m) +
  tm_grid() + 
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_borders() +
  tm_layout(inner.margins = 0)
tmap_arrange(tc1, tc2, tc3)
```

### Raster extraction

```{r 53-raster-extraction-5}
zion_points = read_sf(system.file("vector/zion_points.gpkg", package = "spDataLarge"))
```

```{r 53-raster-extraction-6, purl=FALSE,echo=FALSE}
tm_shape(srtm) +
  tm_raster(style = "cont") +
  tm_shape(zion_points) +
  tm_dots(size = 2)
```

```{r 53-raster-extraction-7}
zion_extract = terra::extract(srtm, vect(zion_points))
zion_points = cbind(zion_points, zion_extract)
zion_points
```

```{r 53-raster-extraction-12}
zion = read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
zion = st_transform(zion, crs(srtm))
zion_srtm_values = terra::extract(srtm, vect(zion))
```

```{r 06-manipulating-raster-data-41}
head(zion_srtm_values)
```

```{r 53-raster-extraction-13, purl=FALSE,echo=FALSE}
tm_shape(srtm) +
  tm_raster(style = "cont") +
  tm_shape(zion) +
  tm_polygons()
```

```{r 53-raster-extraction-14}
zion_srtm_values = terra::extract(srtm, vect(zion))
```

```{r 53-raster-extraction-15, message=FALSE}
library(dplyr)
zion_srtm_values %>% 
  group_by(ID) %>% 
  summarize(across(srtm, list(min = min, mean = mean, max = max)))
```

<!-- exact = TRUE -->

## Raster writing

- https://geocompr.robinlovelace.net/read-write.html#data-output
- https://gdal.org/drivers/raster/gtiff.html

```{r 06-manipulating-raster-data-42, eval=FALSE}
writeRaster(nlcd, filename = "nlcd1.tif", gdal = c("COMPRESS=NONE"))
writeRaster(nlcd, filename = "nlcd2.tif", datatype = "INT1U")
writeRaster(nlcd, filename = "nlcd3.tif", filetype = "COG")
```

## Raster analysis

### Predictions

```{r 06-manipulating-raster-data-43}
landsat_path = system.file("raster/landsat.tif", package = "spDataLarge")
landsat = rast(landsat_path)
zion_points_utm = zion_points_utm = st_transform(zion_points, crs = crs(landsat))
landsat_s = stretch(landsat, maxq = 0.98)
plotRGB(landsat_s, r = 3, g = 2, b = 1)
plot(st_geometry(zion_points_utm), add = TRUE, col = "red", cex = 3)
```

`?predict` - `glm`, `randomForest`, `prcomp`

```{r 06-manipulating-raster-data-44}
zion_points_utm_v = extract(landsat_s, vect(zion_points_utm))
pca = prcomp(zion_points_utm_v[-1])
pca
```

`?predict` - `glm`, `randomForest`, `prcomp`

```{r 06-manipulating-raster-data-45}
pca_pred = predict(landsat_s, pca)
plot(pca_pred)
```

### Interpolations

`?interpolate`

```{r 06-manipulating-raster-data-46}
zion_points_srtm = extract(srtm_utm, 
                      vect(zion_points_utm))
```

```{r 06-manipulating-raster-data-47, message=FALSE}
library(fields)
tps = Tps(st_coordinates(zion_points_utm),
          zion_points_utm$srtm)
rt = rast(srtm_utm)
interp1 = interpolate(rt, tps)
```

```{r 06-manipulating-raster-data-48}
plot(interp1)
```

```{r 06-manipulating-raster-data-49}
library(gstat)
interpolate_gstat = function(model, x, crs, ...) {
	v = st_as_sf(x, coords = c("x", "y"), crs = crs)
	p = predict(model, v, ...)
	as.data.frame(p)[, 1:2]
}
```

```{r 06-manipulating-raster-data-50, warning=FALSE}
v = variogram(srtm ~ 1, data = zion_points_utm)
# plot(v)
mv = fit.variogram(v, vgm(120000, "Exp", 12000, nugget = 10000))
```

```{r 06-manipulating-raster-data-51}
plot(v, model = mv)
```

```{r 06-manipulating-raster-data-52, cache = TRUE, warning=FALSE}
g_OK = gstat(NULL, "srtm", srtm ~ 1, zion_points_utm, model = mv)
OK = interpolate(rt, g_OK, debug.level = 0, fun = interpolate_gstat,
                 crs = crs(rt), index = 1)
plot(OK)
```

### Segmentations

https://github.com/Nowosad/supercells

```{r 06-manipulating-raster-data-53}
library(supercells)
ortho = rast(system.file("raster/ortho.tif", package = "supercells"))
plot(ortho)
```

```{r 06-manipulating-raster-data-54, message=FALSE, results='hide'}
ortho_slic1 = supercells(ortho, k = 200, compactness = 10)
plot(ortho)
plot(st_geometry(ortho_slic1), add = TRUE)
```

```{r 06-manipulating-raster-data-55}
rgb_to_hex = function(x){
  apply(t(x), 2, function(x) rgb(x[1], x[2], x[3], maxColorValue = 255))
}
avg_colors = rgb_to_hex(st_drop_geometry(ortho_slic1[4:6]))

plot(ortho)
plot(st_geometry(ortho_slic1), add = TRUE, col = avg_colors)
```

## Exercises
